alloscore, ISF 43

### 2023-06-24 talks/ISF_talk/talk.qmd
finish alloscore run, work on plot functions for ISF
more plots, working on iter_plot

---
title: "alloscore, ISF 43"
format:
  revealjs: 
    width: 1500
revealjs-plugins:
  - attribution
editor: visual
---

## {.nostretch}

![hubs](hub_timeline.png){}

@@

## The Backstory

Starting \~ 2010? Outbreak forecasting hubs developed to "inform public health responses"

Trend toward elicitation of probabilistic forecasts

-   uncertainty quantification often essential to a public health **DM** (decision maker)

@@

::: columns
::: {.column width="40%"}
Basic element of any such response is an *allocation decision* regarding

@@

::: {.column width="60%"}
Similar to other contexts where there are large-scale institutional efforts to use expert judgment to assess and manage social risk

@@

Even with complex game-theoretic approachs to collective decision making, *allocation* almost always a requirement to strategy development
:::

@@

Hubs have tried to use SRs to rank and combine forecasts in ways that offer best suppport for PH responders

This can be cast in decision-theoretic framework of utility, loss functions, and Bayes acts.

The DM needs to decide take an action $a$ wrt an outcome $Y$

-   such as the number of severe cases in various locations.

Suppose the DM also has a **loss function** $l(a,y)$

-   unmet need when $a < y$ beds are allocated and $y$ severe cases occur

Then the *Bayes act* relative to a forecast $Y \sim F$ is the action $a^F$ that will minimize the DM's expected loss assuming that in reality $Y \sim F$.

@@

this must somehow acknowledge risk attitude heterogeneity among various DM's that a forecast hub might serve
:::

@@

A Scoring Rule attempts to transfer or "share" this loss function with the forecaster through a numerical reward system where FP receives

$$
S(F,y)
$$

units of money, goodwill, prestige, etc. A central goal in designing $S$ is to create a scenario where the FP beleives she cam maximizes her own expected utility by giving the DM the forecast that the FP believes in actually true and/or best suited for leading the DM toward actions that will maximize their utility. An SR that acheieves this is called *proper* and Bayes acts form a linchpin of their operation since the establish an explicit vacabulary of loss in place of vague efforts to rank preformance of various statisticla funcitonals adoptic form statistic inference

## Loss vs Utility

Change in utility but not in loss from change in - supply and personnel cost structures - public opinion regarding importance of fatality prevention - knowledge of demographic disparities in consequences of outcomes

@@

Adopt an iterative Lagrange multiplier method following long history in inventory management of solving *constrained multi-product newsvendor* problems

@@

@@

>>>>

---
title: "alloscore, ISF 43"
format:
  revealjs: 
    width: 1500
revealjs-plugins:
  - attribution
editor: visual
css: styles.css
---

## Evolution of ID Forecasting Hubs {.nostretch}

![](hub_timeline.png){}

@@

## The Role of IDF Hubs

Outbreak forecasting hubs developed to **"inform public health responses"**

- with trend toward elicitation of probabilistic forecasts

- uncertainty quantification essential to a public health **DM** (decision maker)

@@

Basic element of any such response is an **allocation decision** regarding

::: columns

::: {.column width="10%"}

<div style="height: 200px; width: 50px;">
  <svg height="100%" width="100%" viewBox="0 0 10 20">
    <line x1="5" y1="0" x2="5" y2="15" style="stroke:black;stroke-width:2" />
    <polygon points="0,20 10,20 5,15" style="fill:black;" />
  </svg>
</div>

:::

::: {.column width="40%"}

@@

::: {.column width="50%"}
Similar to other contexts where there are large-scale institutional efforts to use expert judgment to assess and manage social risk

@@

Even with complex game-theoretic approachs to collective decision making, allocation almost always a requirement to strategy development
:::

@@

*Very* briefly, a **scoring rule** $S$ assigns a number $S(F,y)$ to the forecast $F$ of $Y$ when $Y=y$ is realized.

$S$ if **proper** if $\displaystyle{E_{F}[S(F,Y)] \leq E_{G}S(F,Y)]}$ for all $F, G$

- ensures that forecasters must give honest forecasts to maximize expected score

Hubs have tried to use SRs to rank and combine forecasts in ways that offer best suppport for PH response to ID outbreaks

Primary scoring rule used currently is the **Weighted Interval Score** which is a discrete approximation to the 
**Continuous Ranked Probability Score**

- adopted for convenient scoring of quantile forecasts.

## Bayes Acts

Scoring rules can be cast in decision-theoretic framework of **Bayes acts**.

::: {.important}

The **Bayes act** relative to a forecast $Y \sim F$ is the action $x^F$ 
by which a DM will minimize **expected loss** $E_Y[l]$ assuming that in reality $Y \sim F$.

:::

 - loss $l$ is a function of the outcome $y$ **and** the action $x$ DM takes in anticipation of $y$

Given $l$, we have an *automatically proper* scoring rule $S(F,y) := l(x^F, y)$.

::: {.callout-note appearance="simple" icon=false}

### Our running example:

 - $x$ and $y$ are vectors in $\mathbb{N}_0^{52}$
 - $y =$ number of severe cases in US states and territories
 - $l(x,y) =$ unmet need when $x < y$ beds are allocated and $y$ severe cases occur

:::

## Formally

$$
\begin{align}
l(x,y) &= \sum_{i=1}^{52} \max(y_i - x_i, 0) \\
x^F &\in \left\{ \text{possible } x' \,\middle| \, E_{F}[l(x',Y)] \leq E_{F}[l(x,Y)] \text{ for any possible } x \right\}
\end{align}
$$

Novelty of our approach for ID forecasting is how we define $x$ being 
**possible**

 - as satisfying some resource constraint $\displaystyle{\sum x_i \leq K}$

::: {.callout-note appearance="simple" icon=false}

### Note:

Our loss function and (single) constraint are very simple prototypes meant 
for proof of concept.

:::

@@

this must somehow acknowledge risk attitude heterogeneity among various DM's that a forecast hub might serve

gpl parameters, weights, multiple constraints
:::

@@

A Scoring Rule attempts to "share" DM loss with the forecaster through $S(F,y)$

- idea goes back to Savage

- may be in units of money, goodwill, prestige

::: {.callout-note appearance="Important" icon=false}

## A central goal in design of scoring rules for social good

Create scenario where the forecaster believes maximizing her own expected utility aligns 
with leading the DM toward actions that maximize social utility.

:::

Using an explicit vacabulary of social loss in ID forecasting hub activities has instrinsic advantage
over use of "off-the-shelf" statistical functionals from statistic inference

::: notes
$S$ a penalty for us

inference has own decision theory and ideas of utility, loss, Bayes acts
that are related but distinct from the fundamentally econommic/policy focussed
ones we use here
:::

## A Lagrangian formulation for the Bayes act

Given $F = \{F_i\}_{i=1}^{N}$, obtaining $x^F$ is a constrained stochastic optimization problem

$$ 
\begin{align}
L(x, \lambda ; K, F) &= \sum_{i=1}^N E_{F_i}\left[l(x_i, Y_i)]\right)+\lambda\left(\sum_{i=1}^N x_i-K\right) \\  
\nabla_{x} L = 0 &\implies 
E_F \left[\frac{d}{dx_i}(Y_i - x_i)_{+}\right] + \lambda 
= F_i(x_i)-1 + \lambda = 0 \\
&\implies x_i = F_i^{-1}(1-\lambda)
\end{align}
$$

Solve $\sum_{l=1}^{N} F_{i}^{-1}(1-\lambda) = K$ to get $x_{i}^F = F_i^{-1}(1-\lambda^{\star})$

@@

Even for simple parametric $F_i$, this rarely has an analytic solution.

Given quantile forecasts $q_{i,j}, j = 1,\ldots,23$, we interpolate full $F_i$ to form $L(x, \lambda ; K, F)$

Adopt an iterative Lagrange multiplier method following long history in inventory management of solving **constrained multi-product newsvendor** problems

@@

## Application

December 2021: Omicron wave clearly started US but forecast teams unsure of severity given uncertainty about $R_0$, cross-protection by vaccination, previous infection, etc.

___

![](plots/state_hosps_ens_only.jpeg)

@@

<!-- ## Loss vs Utility

Change in utility but not in loss from change in - supply and personnel cost structures - public opinion regarding importance of fatality prevention - knowledge of demographic disparities in consequences of outcomes -->
